---
title: "madi_map"
format: html
execute: 
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
# Load libraries
library(tidyverse)
library(sf)
library(terra)
library(biscale)

#Load libraries
library(here)
library(tidyverse)
library(sf)
library(tmap)
library(viridisLite)
library(biscale)
library(lubridate)
library(janitor)
library(readr)
library(ggspatial)
library(ggplot2)
library(cowplot)
```

```{r}
# Global Map Options
# US state boundary data for use as a bounding box
us <- read_sf(here("data", "tl_2024_state", "tl_2024_us_state.shp")) %>% 
  filter(!STUSPS %in% c("MP", "GU", "AS", "AK", "HI", "PR", "VI"))

# make a bounding box
bbox <- st_bbox(us)

# load US counties shapefile
counties <- read_sf(here("data", "tl_2023_us_county", "tl_2023_us_county.shp")) %>% 
  mutate(county = NAME)

# Define  color palettes
wns_palette <- c("white", "blue")
pesticide_palette <- c("white", "red")  
```

#### White-Nose Syndrome and *Pseudomycan destructans* Detections by County

```{r}
# load WNS status by county data
wns <- read_sf(here("data", "wns_county_status", "wns_county_status.csv"))

# join them
wns_counties <- left_join(counties, wns, by = "county") %>% 
  mutate(wns_status = factor(determination, levels = c("Pd Presumed", "WNS Suspect", "Pd Positive", "WNS Positive"), ordered = TRUE)) %>% 
  mutate(date = mdy(release_date)) %>% 
  filter(date <= as.Date("2020-01-01"))

# Exploratory WNS-by-County Map
map_wns_counties <- tm_shape(wns_counties, bbox = bbox)+
  tm_polygons(col = "determination",
              palette = wns_palette,
              NA.col = "white",
              title = "Pd / WNS Status")+
  tm_compass(position = c("left", "bottom"))+
  tm_scale_bar(position = c("left", "bottom"))+
  tm_layout(title = "White-Nose Syndrome Occurrence by County",
            title.position = c("left", "bottom"),
            legend.position = c("right", "bottom"))

# tmap_save(map_wns_counties, here("outputs", "map_wns_counties.png"))
```

![*Pseudomycan destructans* and White-Nose Syndrome Detections by County. Data Source: [U.S. Geological Survey (USGS) National Wildlife Health Center](https://www.usgs.gov/centers/nwhc/science/white-nose-syndrome)](outputs/map_wns_counties.png)

#### Pesticide Use by County

```{r}
pesticides_2013_2017 <- read_csv(here("data", "2013_2017_pesticides", "EPest_county_estimates_2013_2017_v2.csv")) %>% 
  mutate(COUNTYFP = as.character(COUNTY_FIPS_CODE)) %>% 
  mutate(concentration = ifelse(is.na(EPEST_HIGH_KG), 
                                EPEST_LOW_KG, EPEST_HIGH_KG)) %>% 
  group_by(COUNTYFP, YEAR) %>% 
  summarise(total_pesticides_kg = sum(concentration, na.rm = TRUE))  

pesticides_2018 <- read_csv(here("data", "2018_pesticides", "EPest_county_estimates_2018.csv")) %>% 
  mutate(COUNTYFP = as.character(COUNTY_FIPS_CODE)) %>% 
  mutate(concentration = ifelse(is.na(EPEST_HIGH_KG), 
                                EPEST_LOW_KG, EPEST_HIGH_KG)) %>% 
  group_by(COUNTYFP, YEAR) %>% 
  summarise(total_pesticides_kg = sum(concentration, na.rm = TRUE)) 

pesticides_2019 <- read_csv(here("data", "2019_pesticides", "EPest_county_estimates_2019.csv")) %>% 
  mutate(COUNTYFP = as.character(COUNTY_FIPS_CODE)) %>% 
  mutate(concentration = ifelse(is.na(EPEST_HIGH_KG), 
                                EPEST_LOW_KG, EPEST_HIGH_KG)) %>% 
  group_by(COUNTYFP, YEAR) %>% 
  summarise(total_pesticides_kg = sum(concentration, na.rm = TRUE)) 

# join pesticides data
pesticides <- bind_rows(pesticides_2013_2017, pesticides_2018, pesticides_2019)


# join to counties shp
pesticides_counties <- left_join(counties, pesticides, by = "COUNTYFP") %>% 
  mutate(pesticides_per_sqm = total_pesticides_kg/ALAND) %>% 
  mutate(pesticides_log = log(pesticides_per_sqm))


map_pesticides_counties <- tm_shape(pesticides_counties, bbox = bbox)+
  tm_polygons(col = "pesticides_log",
              palette = pesticide_palette,
              breaks = quantile(pesticides_counties$total_pesticides_kg,
                                probs = seq(0, 1, by = 0.10), na.rm = TRUE),
              title = "Total Pesticide Use (Kg)",
              NA.col = "white",
              labels = c("0 - 62 K", "62 - 518 K", "518 K - 1.3 mil", "1.3 - 1.7 mil", 
                         "1.7 - 2 mil", "2 - 2.2 mil", "2.2 - 2.4 mil", 
                         "2.4 - 2.6 mil", "2.6 - 3.1 mil","3.1 - 5.9 mil"))+
  tm_compass(position = c("left", "bottom"))+
  tm_scale_bar(position = c("left", "bottom"))+
  tm_layout(title = "Pesticide Use by County",
            title.position = c("left", "bottom"),
            legend.position = c("right", "bottom"))

# tmap_save(map_pesticides_counties, here("outputs", "map_pesticides_counties.png"))
```

![Pesticide use (kilograms of pesticide) by County. Data Source: [U.S. Geological Survey (USGS) Science-Base Catalog](https://www.sciencebase.gov/catalog/item/6081a924d34e8564d68661a1)](outputs/map_pesticides_counties.png)

#### WNS Status & Pesticide Use by County

```{r}
wns_pesticides <- left_join(wns_counties, pesticides, by = "COUNTYFP") %>% 
  drop_na(total_pesticides_kg) %>% 
  drop_na(wns_status)  %>% 
  mutate(pesticides_per_sqm = total_pesticides_kg/ALAND) %>% 
  mutate(wns_numeric = as.numeric(wns_status)) %>% 
  mutate(pesticides_quantile = cut(pesticides_per_sqm, 
                                  breaks = quantile(pesticides_per_sqm, 
                                                    probs = seq(0, 1, by = 0.25), na.rm = TRUE),
                                  labels = c("Low", "Medium", "High", "Very High")))

write_sf(wns_pesticides, here("data", "wns_pesticides.shp"))

chart_wns_pesticides <- ggplot(data = wns_pesticides, aes(x = wns_status, y = pesticides_per_sqm))+
  geom_col(col = "purple")+
  labs(x = "White-Nose Syndrome Status",
       y = "Pesticide Use (Kg per Meter^2)",
       title = "Pesticide Use by WNS Status")+
  theme_bw()

ggsave(here("outputs", "chart_wns_pesticides.png"), chart_wns_pesticides,
       width = 6, height = 3)
```

# NEW MAP

```{r}
# Classify the pesticide data into categories
wns_pesticides <- wns_pesticides %>% 
  select(wns_numeric, pesticides_quantile, county, geometry)

bivariate <- bi_class(wns_pesticides, 
                      x = wns_numeric, 
                      y = pesticides_quantile, 
                      style = "quantile",
                      dim = 4) %>% 
  drop_na()
```


```{r}
# Set the color palette for the bivariate map
pallet <- "PurpleOr" 

# Create the legend for the bivariate map
legend <- bi_legend(pal = pallet,   
                    flip_axes = FALSE,
                    rotate_pal = FALSE,
                    dim = 4,
                    xlab = "Pesticides (kg/m²)",
                    ylab = "WNS Status",
                    size = 8) +
  theme(plot.background = element_rect(fill = "#f6f9f3"))
```


```{r}
# Create the bivariate map using ggplot2
map <- ggplot() +
  theme_void(base_size = 14) +  
  xlim(-125, -66) +  
  ylim(24, 49) +  
  
  # Plot the bivariate data with appropriate fill color based on bivariate classes
  geom_sf(data = bivariate, aes(fill = bi_class), 
          color = NA, 
          linewidth = 0.1, 
          show.legend = TRUE) +
  
  # Apply the color palette
  bi_scale_fill(pal = pallet, dim = 4, flip_axes = FALSE, rotate_pal = FALSE) +
  
  # Overlay the US state boundaries
  geom_sf(data = us, fill = NA, color = "black", linewidth = 0.4) +  

  # Overlay the county boundaries
  geom_sf(data = counties, fill = NA, color = "grey60", linewidth = 0.25) +  
  
  # Add a scale bar and north arrow together in the bottom right
  annotation_scale(location = "br", width_hint = 0.15, text_cex = 0.8) +  
  annotation_north_arrow(location = "br", which_north = "true",
                         pad_x = unit(0, "cm"), pad_y = unit(1, "cm"), 
                         height = unit(0.9, "cm"), width = unit(0.6, "cm")) +
  
  # Add labels for the map
  labs(title = "White-Nose Syndrome and Pesticide Use in US Counties",
       caption = "Source: U.S. Geological Survey") +
  
  # Customize the appearance of the title, subtitle, and caption
  theme(plot.title = element_text(hjust = 0.5, 
                                  face = "bold"),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(size = 10, 
                                    hjust = 1, 
                                    face = "italic"),
        legend.position = "none",
        plot.background = element_rect(fill = "#f6f9f3", color = NA)) 


# Combine the map and legend using cowplot
finalPlot <- ggdraw() +
  draw_plot(map, 0, 0, 1, 1) +  
  draw_plot(legend, 0.05, 0.05, 0.28, 0.28) 

# Display the final map with legend
finalPlot
```

